Case-Base Neural Network: survival analysis with time-varying, higher-order interactions

### 2023-04-27 StatPaper/APA/wileyNJD-APA.tex
applied Rob's comments

\received{26 April 2023}
\revised{6 June 2023}
\accepted{6 June 2023}

@@

\title{Case-Base Neural Networks: survival analysis with time-varying,
higher-order interactions}

@@

\author[3]{Robert Sladek}

@@

\address[1]{\orgdiv{Department of Quantitative life sciences,} \orgname{McGill University}, \orgaddress{\state{Quebec}, \country{Canada}}}

@@

\presentaddress{McGill genome center: 740 Dr Penfield Ave, Montreal, Quebec H3A 0G1}

\abstract[Summary]{Neural network-based survival methods can model data-driven covariate interactions.
While these methods have led to better predictive performance than regression-based approaches,
they cannot model both time-varying interactions and complex baseline hazards. To address this, we
propose Case-Base Neural Networks (CBNN) as a new approach that combines the case-base sampling
framework with flexible architectures. Our method naturally accounts for censoring and does not require
method specific hyperparameters. Using a novel sampling scheme and data augmentation, we incorporate
time directly into a feed-forward neural network. CBNN predicts the probability of an event occurring at a
given moment to estimate the hazard function. We compare the performance of CBNN to survival methods
based on regression and neural networks in a simulation and three real data applications. We report two time-dependent
metrics for each model. For a complex simulation, which highlights the ability of CBNN to model both a complex
baseline hazard and time-varying interactions and two real data applications, CBNN outperforms all competitors
on a simulation with a complex baseline hazard and time-varying interaction and two case studies. A final case
study shows competitive performance. We highlight the benefit of combining case-base sampling with deep
learning to provide a simple and flexible modeling framework for data-driven, time-varying interaction modeling
of survival outcomes. An R package is available at https://github.com/Jesse-Islam/cbnn.}

@@

Smooth-in-time accelerated failure time (AFT) models can estimate absolute
risks by modeling the hazard directly through a user-specified baseline hazard
distribution \citep{kleinbaum2012survival}. Analysts use Cox proportional hazards
models more often than AFT models \citep{hanley2009}. This preference causes
analyses to be based on hazard ratios and relative risks rather than on survival
curves and absolute risks \citep{hanley2009}. With AFT models, it may be difficult
to identify an appropriate baseline hazard for common diseases with many
interacting risk factors \citep{royston2002flexible}. In studies where the disease
pathogenesis may change with age, we may see non-proportional hazards
\citep{coradini2000time}. For example, previous studies of breast cancer incidence
have discovered time-varying interactions with covariates of interest, such as tumor
size \citep{coradini2000time}. One approach to provide flexibility in the baseline hazard
involves using the basis of splines on time in the model \citep{royston2002flexible}.
However, regression-based models require prior knowledge of potential time-varying
interactions and their quantitative effects.

Neural networks provide a data-driven approach to approximating interaction terms.
For example, DeepSurv is a neural network-based proportional hazards model that
implements the Cox partial log-likelihood as a custom loss function \citep{katzman2018DeepSurv}.
This results in a stepwise absolute risk curve that cannot accommodate time-varying interactions.
Compared to Cox regression, DeepSurv shows better performance on a real dataset
\citep{katzman2018DeepSurv}. Previous work suggests we may change the loss function
to address non-proportional hazards \citep{faraggi1995neural}.

DeepHit assumes an inverse Gaussian distribution as the baseline hazard \citep{lee2018DeepHit}.
It specifies each follow-up time time of interest in the model and directly estimates survival curves, rather
than deriving a hazard function \citep{lee2018DeepHit}. DeepHit outperformed DeepSurv
\citep{lee2018DeepHit}.

We note that these alternative neural network approaches require custom loss functions
\citep{katzman2018DeepSurv} \citep{lee2018DeepHit}. DeepHit introduces a hyperparameter
weighing its two loss functions (negative log-likelihood and ranking losses) \citep{lee2018DeepHit}.
Regression-based approaches require prior specification of all interaction terms, which makes it
challenging to model covariate effects that change with time. The current neural network models
provide flexibility at the cost of clarity, while regression models provide clarity at the cost of flexibility.

In this article, we propose Case-Base Neural Networks (CBNN) as a method that models time-varying
interactions and a flexible baseline hazard. Our approach to modeling the full hazard uses case-base
sampling \citep{hanley2009}. This sampling technique allows probabilistic models to predict survival
outcomes. After case-base sampling, we can easily implement the model in any language using
common neural network components. As part of the case-base framework, we use transformations
of time as a feature (covariate) to specify different baseline hazards. For example, by including
splines of time as covariates, we can approximate the Royston-Parmar flexible baseline hazard model
\citep{royston2002flexible}\citep{hanley2009}. However, this still requires explicit use of time-varying
interactions. CBNN can model both without extra tuning parameters.

We describe how case-base sampling and neural networks combine
in Section \ref{methods}, along with our hyperparameter selection procedure, metrics and software implementation.
We explore the performance of CBNN, DeepSurv, DeepHit, Cox regression and case-base using logistic
regression (CBLR) on simulated data in Section \ref{sims}. Section \ref{casestudies} describes the
performance of the same methods in three case studies. Section \ref{discussion} explores the
implications of our results and contextualizes them within neural network survival analysis in a
single event setting.

@@

In this section, we define case-base sampling, which converts the total
survival time into discrete person-specific moments (person-moments).
Next, we detail how neural networks can be used within this framework,
explicitly incorporating time as a feature while adjusting for the
sampling bias. Then, we describe the metrics and hyperparameter selection procedure.
Finally, we report on the software versions used. An R
package is available for use at
\url{https://github.com/Jesse-Islam/cbnn}. The entire code base to
reproduce the figures and empirical results in this paper is available
at \url{https://github.com/Jesse-Islam/cbnnManuscript}.

@@

Case-base sampling is an alternative framework for survival analysis
\citep{hanley2009}. Without this preliminary step, we cannot model survival
outcomes with probabilistic methods. In case-base sampling, we sample from the continuous
survival time of each person in our dataset to create a \emph{base
series} of \emph{person-moments}. This \emph{base series} complements
the \emph{case series}, which contains all person-moments at which the
event of interest occurs.

@@

Now, let \(b\) be the (user-defined) size of the \emph{base series} and
let \(B\) be the sum of all follow-up times for the individuals in the
study. let $c$ be the number of events in the \emph{case series}.
A reasonable concern is determining how large $b$ should be relative to $c$.
The size of \(b\) influences how much information is lost in the sampling process
\citep{hanley2009}. Using our notation, Hanley $\&$ Miettinen (2009)
adapt a quote from Mantel (1973): "By the reasoning that $\frac{cb}{c+b} [=(\frac{1}{c}+\frac{1}{b}]$
measures the relative information in a comparison of two averages based on
sample sizes of c and b respectively, we might expect by analogy, which would
of course not be exact in the present case, that this approach would result in
only a moderate loss of information. (The practicing statistician is generally
aware of this kind of thing. There is little to be gained by letting the size of
one series, b, become arbitrarily large if the size of the other series, c, must
remain fixed)" \citep{hanley2009} \citep{mantel1}. If $b=100c$, then we expect learned
weights to be at most one percent higher than if the entire study base $B$ was used,
as they are proportional to $\frac{1}{c}+\frac{1}{100c}$ rather than $\frac{1}{c}+\frac{1}{\infty c}$ \citep{hanley2009}.

If we sample the \emph{base series} uniformly across the study
base, then the hazard function of the sampling process is equal to
\(b/B\). Therefore, we have the following equality \cite{saarela2015}
\footnote{We are abusing notation here, conflating hazards with probabilities. For a rigorous treatment, see Saarela \& Hanley (2015) section 3 \cite{saarela2015} .}:
\begin{align}\label{eqn:main}
\frac{P\left(Y_i=1 \mid X_i, T_i\right)}{P\left(Y_i = 0 \mid X_i, T_i\right)} = \frac{h\left(T_i \mid X_i\right)}{b/B}.
\end{align} The odds of a person-moment being a part of the \emph{case
series} is the ratio of the hazard \(h(T_i \mid X_i)\) and the uniform
rate \(b/B\). Using \eqref{eqn:main}, we can see how the log-hazard
function can be estimated from the log-odds arising from case-base
sampling: \begin{align}\label{eqn:offset}
\log \left( h\left(t \mid X_i\right)\right) = \log \left(\frac{P\left(Y_i = 1 \mid X_i, t\right)}{P\left(Y_i = 0 \mid X_i, t\right)}\right) + \log\left(\frac{b}{B}\right).
\end{align}

To estimate the correct hazard function, we adjust for the bias
introduced when sampling a fraction of the study base (\(\frac{b}{B}\)) by adding
the term \(\log\left(\frac{B}{b} \right)\) to offset \(\log\left(\frac{b}{B} \right)\)
during the fitting process. Next, we propose using neural networks to model the
odds.

@@

After case-base sampling, we pass all features, including time, into any
user-defined feed-forward component, to which an offset term is added,
then passed through a sigmoid activation function (Figure
\ref{fig:NNarch}). We are interested in predicting the odds. This makes the sigmoid
activation function ideal as it is the inverse of the odds. which we can use to calculate the hazard. The general form for the
neural network using CBNN is:

@@

where \(T\) is a random variable representing the event time, \(X\) is
the random variable for a covariate profile, \(f_{\theta}(X, T)\)
represents any feed-forward neural network architecture,
\(\log\left(\frac{B}{b}\right)\) is the offset term for adjust for the bias
(\(\log\left(\frac{b}{B}\right)\)) set by case-base sampling,
\(\theta\) is the set of parameters learned by the neural network and
\(\mathrm{sigmoid}(x)=\frac{1}{1+e^{-x}}\). By approximating a
higher-order polynomial of time using a neural network, the baseline
hazard specification is now data-driven, where user-defined
hyperparameters such as regularization, number of layers and nodes
control the flexibility of the hazard function. We provide a detailed
description of the choices we made in the next sub-section.

@@

{\centering \includegraphics[width=1\linewidth]{../figures/nnarch2}}

\caption{Steps involved in CBNN from case-base sampling to the model framework we use for training. The first step is case-base sampling, completed before training begins. Next, we pass this sampled data through a feed-forward neural network. We add the offset and pass that through a sigmoid activation function, whose output is a probability. Once the neural network model completes its training, we can convert the probability output to a hazard, using it for our survival outcomes of interest.}\label{fig:NNarch}
\end{figure}

@@

Backpropagation with an appropriate minimization algorithm (e.g.~Adam,
RMSPropagation, stochastic gradient descent) is used to optimize the
parameters in the model \citep{gulli2017}. For our analysis, we use Adam
as implemented in Keras \citep{gulli2017}. Note that the size of the
\emph{case series} is fixed as the number of events, but we can make the
\emph{base series} as large as we want. A ratio of 100:1 \emph{base
series} to \emph{case series} is sufficient \citep{hanley2009}. We pass
our feed-forward neural network through a sigmoid activation function
(Figure \ref{fig:NNarch}). Finally, we can convert this model output to
a hazard. To use CBNN for predictions, we manually set the
offset term \(\log\left(\frac{B}{b} \right)\) to 0 in the new data as we already account for the bias during the
fitting process.

@@

To choose our method-specific hyperparameters, we use the Integrated
Brier Score (IBS) \citep{graf1999}. We use two metrics to assess the
performance of the different methods of interest on testing data: 1) Index of Prediction Accuracy (IPA)
\citep{kattan2018index} and 2) Inverse probability censoring weights-adjusted time dependent
area under the receiver operating characteristic curve ($AUC_{IPCW}$) \citep{auc}, which we 
define below. We achieve a summarized assessment of performance for each model with the
IBS. The IPA is a metric for both discrimination and calibration. The $AUC_{IPCW}$
provides a metric for discrimination only. Together, we can get a clearer picture about how
each model performs in each study and why.

@@

The IPA is a function of the Brier score (\(BS(t)\)) \citep{graf1999},
which is defined as \begin{align}\label{eqn:bs}
BS(t)=\frac{1}{N}\sum^{N}_{i=1}\left(\frac{\left(1 - \widehat{F}(t \mid X_{i})\right)^{2}\cdot I(T_{i}\leq t,\delta_{i}=1)}{\widehat{G}(T_{i})} + \frac{\widehat{F}(t\mid X_{i})^{2}\cdot I(T_{i}>t)}{\widehat{G}(t)}\right),
\end{align} where \(\delta_{i}=1\) shows individuals who have
experienced the event, \(N\) represents the number of samples in our
dataset over which we calculate \(BS(t)\), \(\widehat{G}(t)=P[c>t]\) is
a non-parametric estimate of the censoring distribution, \(c\) is
censoring time and \(T_{i}\) is an individual's survival or censoring
time. The BS provides a score that accounts for the information
loss because of censoring. There are three categories of individuals
that may appear within the dataset once we fix our \(t\) of interest.
Individuals who experienced the event before \(t\) are present in the
first term of the equation. The second term of the equation includes
individuals who experience the event or are censored after \(t\). Those
censored before \(t\) are the third category of people. The inverse
probability censoring weights (IPCW) adjustment (\(G(\cdot)\)) is to
account for these category three individuals whose information is
missing.

The IPA as a function of time is given by \begin{align}
\textrm{IPA}(t) &= 1-\frac{BS_{model}(t)}{BS_{null}(t)}, \nonumber
\end{align} where \(BS_{model}(t)\) represents the BS over time
\(t\) for the model of interest and \(BS_{null}(t)\) represents the
BS if we use an unadjusted Kaplan-Meier (KM) curve as the
prediction for all observations \citep{kattan2018index}. Note that IPA
has an upper bound of one, where positive values show an increase in
performance over the null model and negative values show that the null
model performs better. These scores show how performance changes over
follow-up time. As an extension of BS, IPA models for both calibration
and discrimination\citep{graf1999} \citep{kattan2018index}.

The Integrated Brier Score (IBS) is a function of the BS \ref{eqn:bs} as well \citep{graf1999},
which is defined as \begin{align*}
IBS(t)=\int_{t}^{t_{max}}BS(t)dw(t),
\end{align*} where $w(t)=\frac{t}{t_{max}}$. As this is a score for a range of follow-up-times,
it is a useful metric to track during cross-validation. The performance at a specific time is lost,
and is why we opt for the IPA in assessing performance on a test set.

@@

The IPCW-adjusted AUC ($AUC_{IPCW}$) is a time-dependent metric that considers censoring \citep{auc}. 
For a given follow-up time of interest,
 \begin{align*}
AUC_{IPCW}(t)=\frac{\sum^{n}_{i=1}\sum^{n}_{j=1} I \left(\widehat{F}(t|X_{i})>\widehat{F}(t|X_{j})\right) \cdot I(T_{i}\leq t,\delta_{i}=1) \cdot \left( 1-I(T_{j}\leq t,\delta_{j}=1) \right) \cdot W_{i}(t) \cdot W_{j}(t)} { \sum^{n}_{i=1}\sum^{n}_{j=1} I(T_{i}\leq t,\delta_{i}=1) \cdot \left( 1-I(T_{j}\leq t,\delta_{j}=1) \right) \cdot W_{i}(t) \cdot W_{j}(t) },
\end{align*}where \begin{align*}
W_{i}(t)=\frac{I(T_{i}\leq t,\delta_{i}=1)}{\widehat{G}(T_{i})} +\frac{I(T_{i}>t)}{\widehat{G}(t)}.
\end{align*} The $AUC_{IPCW}$ measures discrimination \citep{auc}.
By taking both IPA and $AUC_{IPCW}$ together, we can better understand whether a model is better
in terms of calibration or discrimination.

@@

Neural networks are flexible when defining the architecture and optimization parameters.
These hyperparameter choices can affect the estimated parameters and are not necessarily
the same for each model. We apply the same procedure to each study in this paper. First,
we fix a test set with 15\% of the data. We keep aside this data during hyperparameter
selection and we fit a model for each method in each study. To determine the best
hyperparameters for each neural network method, we use three-fold cross-validated grid
search \citep{gulli2017} on the remaining data (85\% training, 15\% validation) for the
following range of hyperparameters:

\[
Grid\ search: \begin{cases}
Learning\ rate \sim \{0.001, 0.01\} \\ %technically a control and treatement group
Dropout \sim \{0.01,0.05, 0.1\} \\
First\ layer\ nodes \sim \{50, 75, 100\} \\
Second\ layer\ nodes \sim \{10,25,50\} \\
Number\ of\ batches \sim \{100, 500\} \\
Activation\ function \sim \{ReLU, Linear\}\\
\alpha \sim \{0, 0.5, 1\}.
\end{cases}
\]

\begin{table}
\caption{Four tables showing the chosen hyperparameters along with the IBS for each neural network model in the complex simulation (A), multiple myeloma (MM) case study (B), free light chain (FLC) case study (C) and prostate cancer (Prostate) case study (D).}
\label{tab:wins}
\begin{center}\includegraphics[width=1\linewidth]{../figures/winTable} \end{center}
\end{table}

DeepHit uses $\alpha$ as a specific hyperparameter, while DeepSurv and CBNN do not.
We track IBS on the validation set for each hyperparameter combination,
choosing the combination with the lowest score for each method.
We present the chosen hyperparameters in Table 1.

@@

R \citep{Rsoft} and python \citep{py} are used to evaluate methods from
both languages. We fit the Cox model using the \textbf{survival} package
\citep{survpkg}, the CBLR model using the \textbf{casebase} package
\citep{cbpkg}, DeepSurv and DeepHit using \textbf{pycox}
\citep{lee2018DeepHit}. We made the components
of CBNN using the \textbf{casebase} package for the sampling step and
the \textbf{keras} \citep{keras} package for our neural network
architecture. The \textbf{simsurv} package \citep{simsurv} is used for
our simulation studies, while \textbf{flexsurv} \citep{flexsurv} is used
to fit a flexible baseline hazard using splines for our complex
simulation. The \textbf{riskRegression}
package \citep{riskRegression} is used to get the Index of Prediction
Accuracy (IPA metric) and AUC. Both metrics are described in detail in the
following section. We modify the \textbf{riskRegression} package to be
used with any user supplied risk function \(F\). To ensure that both R
and Python-based models are running in unison on the same data through
our simulations and bootstrap, we use the \textbf{reticulate} package
\citep{reticulate}.

@@

{\centering \includegraphics[width=1\linewidth]{../figures/megaPlot}

@@

We would like to thank Dr.~James Meigs, the project
leader of these awards, for his support and helpful discussions. The
work was also supported as part of the Congressionally Directed Medical
Research Programs (CDMRP) award W81XWH-17-1-0347. We would also like to
thank Dr.~James Hanley for his support and discussions while extending
the case-base methodology.

>>>>

\received{XXXXXX}
\revised{XXXXXX}
\accepted{XXXXXX}

@@

\title{Case-Base Neural Network: survival analysis with time-varying,
higher-order interactions}

@@

\author[1,3]{Robert Sladek}

@@

\address[1]{\orgdiv{Department of Quantitative Life Sciences,} \orgname{McGill University}, \orgaddress{\state{Quebec}, \country{Canada}}}

@@

\presentaddress{McGill Genome Center: 740 Dr Penfield Ave, Montreal, Quebec H3A }

\abstract[Summary]{Neural network-based survival methods can model data-driven covariate interactions. While these methods have led to better predictive performance than regression-based approaches, they cannot model time-varying interactions and complex baseline hazards simultaneously. To address this, we propose Case-Base Neural Network (CBNN) as a new approach that combines the case-base sampling framework with flexible neural network architectures. Using a novel sampling scheme and data augmentation to naturally account for censoring, we construct a feed-forward neural network that may take time as an input. CBNN predicts the probability of an event occurring at a given moment to estimate the hazard function. We compare the performance of CBNN to neural network and regression-based survival methods in a simulation and three case studies using two time-dependent metrics. First, we examine performance on a simulation involving a complex baseline hazard and time-varying interactions to assess all methods, with CBNN outperforming competitors. Then, we apply all methods to two real data applications where CBNN outperforms the competing models as well. In a third case study, CBNN shows competitive performance. Our results highlight the benefit of combining case-base sampling with deep learning to provide a simple and flexible modeling framework for data-driven, time-varying interaction modeling of single event survival outcomes. An R package is available at https://github.com/Jesse-Islam/cbnn.}

@@

A common assumption we use in survival analysis is that the risk of the event of interest does not vary with
time (i.e. proportional hazards) \citep{hanley2009}. This simplifying assumption results in using Cox proportional
hazards models more often than Smooth-in-time, accelerated failure time (AFT) models \citep{hanley2009}. This
preference causes analyses to be based on hazard ratios and relative risks rather than on survival curves and
absolute risks \citep{hanley2009}. The proportional hazards assumption in a Cox model may be incorrect in studies
where the disease pathogenesis may change over time \citep{coradini2000time}. For example, studies of breast cancer
incidence have shown that the risk of developing a tumor involves a time-varying interaction with tumor size
\citep{coradini2000time}. Time-varying covariates can be incorporated easily into AFT models \citep{royston2002flexible}.
To provide flexibility in the baseline hazard, we can use a basis of splines on time as model inputs
\citep{royston2002flexible}. However, these regression-based models require prior knowledge of potential
time-varying interactions and their quantitative effects.

Neural networks provide a data-driven approach to approximating interaction terms in models. For example.
DeepSurv is a neural network-based proportional hazards model that implements the Cox partial log-likelihood
as a custom loss function \citep{katzman2018DeepSurv}. This results in a stepwise absolute risk curve that cannot
accommodate time-varying interactions. Compared to Cox regression, DeepSurv shows better performance on a
real dataset \citep{katzman2018DeepSurv}. Previous work suggests we may change the loss function to address
non-proportional hazards \citep{faraggi1995neural}. In contrast, DeepHit assumes an inverse Gaussian distribution
as the baseline hazard \citep{lee2018DeepHit} and specifies each follow-up time of interest in the model. DeepHit
directly estimates survival curves, rather than deriving a hazard function \citep{lee2018DeepHit}. On real datasets,
DeepHit outperformed DeepSurv \citep{lee2018DeepHit}. We note that DeepSurv and DeepHit both use custom loss
functions \citep{katzman2018DeepSurv} \citep{lee2018DeepHit}. In addition, DeepHit introduces a hyperparameter
weighing its two loss functions (negative log-likelihood and ranking losses) \citep{lee2018DeepHit}. Compared
to these neural network methods, regression-based approaches require prior specification of all interaction terms,
which makes it challenging to model covariate effects that change with time. The current neural network models
provide flexibility at the cost of clarity, while regression models provide clarity at the cost of flexibility.

We propose Case-Base Neural Networks (CBNN) as a method for single event survival analysis that models
time-varying interactions and a flexible baseline hazard. CBNN use the case-base sampling technique, which
allows probabilistic models to predict survival outcomes \citep{hanley2009}. After case-base sampling, we
implement a model using common neural network components that uses transformations of time as a feature
(covariate) to specify different baseline hazards. For example, by including splines of time as covariates,
case-base with logistic regression (CBLR) can approximate the Royston-Parmar flexible baseline hazard
\citep{royston2002flexible} \citep{hanley2009}. However, this still requires explicit specification of
time-varying interactions. CBNN can model both without extra tuning parameters.

In this paper, we describe the case-base sampling method and compare its properties to neural
network models in Section \ref{methods}, along with our hyperparameter selection procedure,
metrics and software implementation. We also explore the performance of CBNN, DeepSurv,
DeepHit, Cox regression and CBLR on simulated data in Section \ref{sims}. Section \ref{casestudies}
describes the performance of the same methods in three case studies. Section \ref{discussion} explores
the implications of our results and contextualizes them within neural network survival analysis in a single
event setting.

@@

Case-base sampling divides total survival time for each individual into person-specific moments
(person-moments), treating each person-moment as a sample. In this section, we detail how neural
networks explicitly incorporating time as a feature while adjusting for the sampling bias in case-base
sampled data; and describe the metrics and hyperparameter selection procedure used to compare
CBNN with other software packages. An R package to use CBNN is available at
\url{https://github.com/Jesse-Islam/cbnn}. The entire code base to reproduce the figures
and empirical results in this paper is available at \url{https://github.com/Jesse-Islam/cbnnManuscript}.

@@

Case-base sampling is an alternative framework for survival analysis \citep{hanley2009},
which converts the total survival time into discrete person-moments. Without this preliminary step,
we cannot model survival outcomes with probabilistic methods. In case-base sampling, we sample
from the continuous survival time of each person in our dataset to create a \emph{base series} of
\emph{person-moments}. This \emph{base series} complements the \emph{case series}, which contains
all person-moments at which the event of interest occurs.

@@

Now, let \(b\) be the (user-defined) size of the \emph{base series} and let \(B\) be the sum of all follow-up
times for the individuals in the study. Let $c$ be the number of events in the \emph{case series}. A
reasonable concern is determining how large $b$ should be relative to $c$, since the size of \(b\)
influences how much information is lost in the sampling process \citep{hanley2009}. Using our notation,
Hanley $\&$ Miettinen (2009) adapt a quote from Mantel (1973): "By the reasoning that
$\frac{cb}{c+b}=\left[\left( \frac{1}{c}+\frac{1}{b}\right)]\right]$ measures the relative information in a comparison of two
averages based on sample sizes of c and b respectively, we might expect by analogy, which would
of course not be exact in the present case, that this approach would result in only a moderate loss of
information. (The practicing statistician is generally aware of this kind of thing. There is little to be gained
by letting the size of one series, b, become arbitrarily large if the size of the other series, c, must remain
fixed)" \citep{hanley2009} \citep{mantel1}. If $b=100c$, then we expect learned weights to be at most one
percent higher than if the entire study base $B$ was used, as they are proportional to
$\frac{1}{c}+\frac{1}{100c}$ rather than $\frac{1}{c}+\frac{1}{\infty c}$ \citep{hanley2009}.

If we sample the \emph{base series} uniformly across the study base, then the hazard function
of the sampling process is equal to \(b/B\). Therefore, we have the following equality \cite{saarela2015}
\footnote{For a rigorous treatment, see Saarela \& Hanley (2015) section 3 \cite{saarela2015} .}:
\begin{align}\label{eqn:main}
\frac{P\left(Y_i=1 \mid X_i, T_i\right)}{P\left(Y_i = 0 \mid X_i, T_i\right)} = \frac{h\left(T_i \mid X_i\right)}{b/B}.
\end{align} The odds of a person-moment being a part of the \emph{case
series} is the ratio of the hazard \(h(T_i \mid X_i)\) and the uniform
rate \(b/B\). Using \eqref{eqn:main}, we can see how the log-hazard
function can be estimated from the log-odds arising from case-base
sampling: \begin{align}\label{eqn:offset}
\log \left( h\left(t \mid X_i\right)\right) = \log \left(\frac{P\left(Y_i = 1 \mid X_i, t\right)}{P\left(Y_i = 0 \mid X_i, t\right)}\right) + \log\left(\frac{b}{B}\right).
\end{align}

To estimate the correct hazard function, we adjust for the bias introduced when sampling a fraction of the
study base (\(\frac{b}{B}\)) by adding the term \(\log\left(\frac{B}{b} \right)\) to offset \(\log\left(\frac{b}{B} \right)\)
during the fitting process.

@@

After case-base sampling, we pass all features, including time, into any user-defined feed-forward
component. Then, we add an offset term and pass the output through a sigmoid activation function
(Figure \ref{fig:NNarch}). Since we are interested in predicting the odds of an event occurring, the sigmoid
activation function is ideal as it is the inverse of the odds and can be used to calculate the hazard. The
general form for the neural network using CBNN is:

@@

where \(T\) is a random variable representing the event time, \(X\) is the random variable for a covariate
profile, \(f_{\theta}(X, T)\) represents any feed-forward neural network architecture, \(\log\left(\frac{B}{b}\right)\)
is the offset term for adjust for the bias (\(\log\left(\frac{b}{B}\right)\)) set by case-base sampling, \(\theta\) is the
set of parameters learned by the neural network and \(\mathrm{sigmoid}(x)=\frac{1}{1+e^{-x}}\). By approximating
a higher-order polynomial of time using a neural network, the baseline hazard specification is now data-driven,
where user-defined hyperparameters such as regularization, number of layers and nodes control the flexibility
of the hazard function.

@@

{\centering \includegraphics[width=1\linewidth]{../figures/Figure1.pdf}}

\caption{Methodological steps involved in CBNN. The first step, case-base sampling, is completed before training begins.
We pass this sampled data through a feed-forward neural network and add an offset to adjust for the bias inherent in
case-base sampling, then apply a sigmoid activation function to create a probability. Once the neural network model
completes its training, we can convert the probability output to a hazard for the survival outcomes of interest.}\label{fig:NNarch}
\end{figure}

@@

Backpropagation with an appropriate minimization algorithm (e.g.~Adam, RMSPropagation, stochastic gradient
descent) is used to optimize the parameters in the model \citep{gulli2017}. For our analysis, we use Adam as implemented
in Keras \citep{gulli2017}. While the size of the \emph{case series} is fixed as the number of events; the size of the
\emph{base series} is not restricted. We use a ratio of 100:1 \emph{base series} to \emph{case series} \citep{hanley2009}.
We pass our feed-forward neural network through a sigmoid activation function (Figure \ref{fig:NNarch}). Finally, we can
convert this model output to a hazard. To use CBNN for predictions, we manually set the offset term \(\log\left(\frac{B}{b} \right)\)
to 0 in the new data as we already account for the sampling bias during the fitting process.

@@

To choose our method-specific hyperparameters, we used the Integrated Brier Score (IBS) \citep{graf1999}, which is based on the
Brier Score (BS) and provides a summarized assessment of performance for each model. We assess the performance of models on
a held out dataset using two metrics: 1) the Index of Prediction Accuracy (IPA) \citep{kattan2018index}; and 2) the Inverse probability
censoring weights-adjusted time dependent area under the receiver operating characteristic curve ($AUC_{IPCW}$) \citep{auc}. The
IPA score is a metric for both discrimination and calibration, while the $AUC_{IPCW}$ provides a metric for discrimination only.

\hypertarget{bs}{%
\subsubsection{Brier Score (BS)}\label{bs}}
The BS \citep{graf1999} is defined as \begin{align}\label{eqn:bs}
BS(t)=\frac{1}{N}\sum^{N}_{i=1}\left(\frac{\left(1 - \widehat{F}(t \mid X_{i})\right)^{2}\cdot I(T_{i}\leq t,\delta_{i}=1)}{\widehat{G}(T_{i})} + \frac{\widehat{F}(t\mid X_{i})^{2}\cdot I(T_{i}>t)}{\widehat{G}(t)}\right),
\end{align} where \(\delta_{i}=1\) shows individuals who have experienced the event, \(N\) represents the number of samples
in our dataset over which we calculate \(BS(t)\), \(\widehat{G}(t)=P[c>t]\) is a non-parametric estimate of the censoring distribution,
\(c\) is censoring time and \(T_{i}\) is an individual's survival or censoring time. The BS provides a score that accounts for the
information loss because of censoring. Once we fix our \(t\) of interest, the individuals in the dataset can be divided into three
groups. Individuals who experienced the event before \(t\) are present in the first term of the equation. The second term of the
equation includes individuals who experience the event or are censored after \(t\). Those censored before \(t\) (the remaining
individuals) are accounted for by the IPCW adjustment (\(G(\cdot)\)) present on both terms.

The Integrated Brier Score (IBS) is a function of the BS \ref{eqn:bs} \citep{graf1999}, which is defined as \begin{align*}
IBS(t)=\int_{t}^{t_{max}}BS(t)dw(t),
\end{align*} where $w(t)=\frac{t}{t_{max}}$. As the IBS is calculated for a range of follow-up-times,
it is a useful metric to track overall performance across all of follow-up time during cross-validation. We use the
IPA score to get an estimate of performance at each follow-up time for our studies.

@@

The IPA as a function of time, based on the BS. The IPA score is given by \begin{align}
\textrm{IPA}(t) &= 1-\frac{BS_{model}(t)}{BS_{null}(t)}, \nonumber
\end{align} where \(BS_{model}(t)\) represents the BS over time \(t\) for the model of interest and \(BS_{null}(t)\) represents
the BS if we use an unadjusted Kaplan-Meier (KM) curve as the prediction for all observations \citep{kattan2018index}. The
IPA score has an upper bound of one, where positive values show an increase in performance over the null model and
negative values show that the null model performs better. An extension of BS, the IPA scores assess both model calibration
and discrimination and show how performance changes over follow-up time \citep{graf1999} \citep{kattan2018index}.

@@

The IPCW-adjusted AUC ($AUC_{IPCW}$) is a time-dependent metric that considers censoring \citep{auc}. For a given follow-up time of interest,
 \begin{align*}
AUC_{IPCW}(t)=\frac{\sum^{n}_{i=1}\sum^{n}_{j=1} I \left(\widehat{F}(t|X_{i})>\widehat{F}(t|X_{j})\right) \cdot I(T_{i}\leq t,\delta_{i}=1) \cdot \left( 1-I(T_{j}\leq t,\delta_{j}=1) \right) \cdot W_{i}(t) \cdot W_{j}(t)} { \sum^{n}_{i=1}\sum^{n}_{j=1} I(T_{i}\leq t,\delta_{i}=1) \cdot \left( 1-I(T_{j}\leq t,\delta_{j}=1) \right) \cdot W_{i}(t) \cdot W_{j}(t) },
\end{align*}where \begin{align*}
W_{i}(t)=\frac{I(T_{i}\leq t,\delta_{i}=1)}{\widehat{G}(T_{i})} +\frac{I(T_{i}>t)}{\widehat{G}(t)}.
\end{align*} The $AUC_{IPCW}$ measures discrimination \citep{auc}. By examining both IPA and $AUC_{IPCW}$,
we can better understand whether a model performs better in terms of calibration or discrimination.

@@

Neural networks require external parameters that constrain the learning process (hyperparameters). We apply the following hyperparameter
optimization procedure to each method for each study. First, we fix a test set with 15\% of the data which we keep aside during hyperparameter selection.
To determine the best hyperparameters for each neural network method, we use a three-fold cross-validated grid search \citep{gulli2017} on the remaining
data (85\% training, 15\% validation) for the following range of hyperparameters:
\[
Grid\ search: \begin{cases}
Learning\ rate \sim \{0.001, 0.01\} \\ 
Dropout \sim \{0.01,0.05, 0.1\} \\
First\ layer\ nodes \sim \{50, 75, 100\} \\
Second\ layer\ nodes \sim \{10,25,50\} \\
Number\ of\ batches \sim \{100, 500\} \\
Activation\ function \sim \{ReLU, Linear\}\\
\alpha \sim \{0, 0.5, 1\}. \ (DeepHit \ only)
\end{cases}
\]

\begin{table}
\caption{Hyperparameters selected after three-fold cross-validated grid search along with the average IBS for each neural network model in the
complex simulation (A); multiple myeloma (MM) case study (B); free light chain (FLC) case study (C); and prostate cancer (Prostate) case study (D).}
\label{tab:wins}
\begin{center}\includegraphics[width=1\linewidth]{../figures/Table1.pdf} \end{center}
\end{table}

DeepHit uses $\alpha$ as a hyperparameter, while DeepSurv and CBNN do not. We track IBS on the validation set for each hyperparameter combination,
choosing the combination with the lowest score for each method (Table \ref{tab:wins}).

@@

R \citep{Rsoft} and Python \citep{py} are used to evaluate methods from both languages. We fit the Cox model using the \textbf{survival} package
\citep{survpkg} and the CBLR model using the \textbf{casebase} package \citep{cbpkg}. Both DeepSurv and DeepHit are fit using \textbf{pyCox}
\citep{lee2018DeepHit}. We made the components of CBNN using the \textbf{casebase} package \citep{cbpkg} for the sampling step and the
\textbf{keras} \citep{keras} package for our neural network architecture. We use the \textbf{simsurv} package \citep{simsurv} for our simulation studies
and \textbf{flexsurv} \citep{flexsurv} to fit a flexible baseline hazard using splines for our complex simulation. The \textbf{riskRegression} package
\citep{riskRegression} is used to get the IPA and $AUC_{IPCW}$. We modify the \textbf{riskRegression} package to be used with any user supplied
risk function \(F\). We use the \textbf{reticulate} package \citep{reticulate} to run both R based methods and python based methods on the same seed.

@@

{\centering \includegraphics[width=1\linewidth]{../figures/Figure2.pdf}

@@

We would like to thank Dr.~James Meigs, the project
leader of UM1DK078616 and R01HL151855, for his support and helpful discussions. The
work was also supported as part of the Congressionally Directed Medical
Research Programs (CDMRP) award W81XWH-17-1-0347. We would also like to
thank Dr.~James Hanley for his support and discussions while extending
the case-base methodology.
